Data exploration through dot-driven development\footnote{This work was supported by The Alan Turing Institute under the EPSRC grant EP/N510129/1 and by the Google Digital News Initiative.}

### 2017-05-16 ecoop-2017/paper.tex
Final version edits
Final artifact version
Final version

\author[1]{Tomas Petricek}
\affil[1]{The Alan Turing Institute, London, UK\\
  \texttt{tomas@tomasp.net}}
\authorrunning{T. Petricek}
\Copyright{Tomas Petricek}
\subjclass{D.3.2 Very high-level languages}
\keywords{Data science, type providers, pivot tables, aggregation}

\EventEditors{John Q. Open and Joan R. Acces}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{DRAFT}
\EventAcronym{DRAFT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}

@@

The rise of big data and open data initiatives means that there is an increasing amount of raw data 
available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 \cite{posttruth} 
suggests that there has never been a greater need for increasing data literacy and building tools that let 
anyone -- including journalists and interested citizens -- explore such data and use it to make
transparent factual claims.

@@

In Figure~\ref{fig:pivot-members}, we list a subset of the provided members that were used in the
example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
the possible sorting keys and directions. After choosing \qident{group data}, the user first 
selects the grouping key and then can choose one or more aggregations that can be applied on
other columns of the group. Thus an entire data transformation (such as choosing top 8 athletes
by the number of gold medals) can be constructed using dot-driven development.

@@

Our language supports method calls such as $\ident{head}(\num{8})$ and an alternative design for
filtering would be to provide a method such as $\qident{Games is}(\str{Rio (2016)})$. However,
the fact that we can offer possible values largely simplifies writing of the script for the most 
common case when the user is interested in one of the known values.

@@

\noindent
When grouping, we sum the number of gold and silver medals and concatenates team names~\ball{1}. 
Then we sort the grouped data using multiple sorting keys \ball{2} -- first by the number of gold 
medals and then silver medals (within a group with the same number of gold medals).

@@

In type providers for F\#, this could be done using methods with static parameters \cite{staticpar} by writing, 
for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
the static parameter (here, \str{Total Gold}) is passed to the type provider, which can use it to
generate the type signature of the method and the return type with member name according to the
value of the static parameter.

@@

\noindent
Operations that change the type of rows in the data set can be captured using row types~\cite{rowtypes}. 
In addition, we need to annotate type with a form of typestate \cite{typestate} to restrict what operations are available. 
When dropping columns, we first access the \qident{drop columns} member, which sets the state to
a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
then be used to complete the operation and choose another transformation.

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\begin{array}{rcl}
  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (E) \\
    \\
 \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}
\vspace{-1.5em}

\[
\inference[(\emph{member})~]
{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
\]
\vspace{-3em}

@@

The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
models a type-safe data series mapping keys of type $\tau_1$ to values of type $\tau_2$ that
can be used, for example, as input for a charting library. A series is a typed wrapper over a 
\indet{Query} value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained 
from the pivot type provider contains keys and values of matching~types.

@@

\noindent
The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
members of $C_1$ remain unevaluated when they do not appear in the source code. This aspect of type
providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
The typing rules given are written using the following judgement:
%
\begin{equation*}
L_1; \Gamma \vdash e : \tau; L_2
\end{equation*}

@@

\begin{figure}[t]
\vspace{-1.5em}
\[
\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
\quad
\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
\quad
\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
\]
\vspace{-2.25em}

\[
\inference[(\emph{data})~]
  {L; \Gamma \vdash v_{i,j} : \tau;L \quad \tau \in \{ \kvd{num}, \kvd{string} \}}
  {L; \Gamma \vdash \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} : \ident{Query};L }
\]
\vspace{-2.25em}

@@

\[
\inference[(\emph{sel})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \sigma_\varphi (e) : \ident{Query}; L_2}
\qquad
\inference[(\emph{group})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(e): \ident{Query}; L_2}
\]
\vspace{-2.25em}

\[
\inference[(\emph{series})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \ident{series}\langl \tau_1, \tau_2 \rangl(e) : \ident{series}\langl \tau_1, \tau_2 \rangl; L_2}
\]
\vspace{-2.25em}

\[
\inference[(\emph{new})~]
  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots),L}
  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
\]
\vspace{-2.25em}

\[
\inference[(\emph{member})~]
  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2\cup L; \Gamma,x:\tau \vdash e_i : \tau_i; L_3 \\
   L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L }
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_3}
\]
\vspace{-0.5em}
\caption{Type-checking of Foo expressions with lazy context}
\vspace{-0.5em}
\label{fig:foo-typecheck}
\end{figure}

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\begin{array}{ll}
\ident{drop}(F) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\qquad\ball{1}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f = C_f(\Pi_{\ident{dom}(F')}(x)) & \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\\
\ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
\\[0.5em]
\multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
\quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))\quad~~\ball{5}           &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\end{array}
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}

@@

@@

\begin{figure}[t]
\begin{equation*}
\begin{array}{ll}
  \ident{pivot}(F, D) = C, \{ C \mapsto (l, L_1 \cup L_2 \cup \ldots  \}\quad \ball{1}\\[0.5em]
  ~ l = \kvd{type}~C(x:\ident{Query}) = \\
  \quad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F, D)\\
  \quad \qquad \kvd{member}~\qident{filter data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{filter}(F, D)\\
  \quad \qquad (\ldots)\\
\\
\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\quad\ball{2}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{3}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{4}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{5} \\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
\quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{6}
\end{array}
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\hspace{-1em}
\begin{array}{l}
\ident{olympics}.\qident{filter data}.\qident{Medal is}.\ident{Gold}.\qident{Team is}\\[0.25em]
\quad\begin{array}{l}
\rightarrow \qident{Czech Republic}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Barbora Spotakova}\\
\qquad \rightarrow \qident{David Kostelecky}\\
\qquad \rightarrow \qident{David Svoboda}\\
\end{array}\qquad
\begin{array}{l}
\rightarrow \ident{Mongolia}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Badar-Uugan Enkhbat}\\
\qquad \rightarrow \qident{Tuvshinbayar Naidan}\\
\\
\end{array}
\end{array}
\end{equation*}
\caption{Subset of members provided by the filtering operation}
\label{fig:pivot-filter}
\end{figure}

@@

\begin{figure}[b]
\begin{center}
\includegraphics[scale=0.30,trim=0mm 0mm 0mm 0mm,clip]{images/table.png} % left bottom right top
\end{center}
\caption{Athletes by the number of medals over the entire history of Olympic games}
\label{fig:case-table}
\end{figure}

@@

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.31,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
\end{center}
\caption{Spreadsheet-inspired live editor for the pivot type provider}
\label{fig:case-ed}
\end{figure}

>>>>

\author[1]{Tomas Petricek}
\affil[1]{The Alan Turing Institute, London, UK\\
 and Microsoft Research, Cambridge, UK\\
  \texttt{tomas@tomasp.net}}
\authorrunning{T. Petricek}
\Copyright{Tomas Petricek}
\subjclass{D.3.2 Very high-level languages}
\keywords{Data science, type providers, pivot tables, aggregation}

\EventEditors{Peter M\"uller}
\EventNoEds{1}
\EventNoEds{1}
\EventLongTitle{31st European Conference on Object-Oriented Programming (ECOOP 2017)} 
\EventShortTitle{ECOOP 2017} 
\EventAcronym{ECOOP} 
\EventYear{2017} 
\EventDate{June 18--23, 2017} 
\EventLocation{Barcelona, Spain} 
\EventLogo{} 
\SeriesVolume{74} 
\ArticleNo{55}

@@

The rise of big data and open data initiatives means that there is an increasing amount of raw data 
available. At the same time, the fact that ``post-truth'' was chosen as the word of 2016 \cite{posttruth} 
suggests that there has never been a greater need for increasing data literacy and tools that let 
anyone explore such data and use it to make transparent factual claims.

@@

Figure~\ref{fig:pivot-members} lists a subset of the members from the
example in Section~\ref{sec:intro}. After choosing \qident{sort data}, the user is offered 
the possible sorting keys. After choosing \qident{group data}, the user first 
selects the grouping key and then can choose one or more aggregations that can be applied on
other columns of the group. Thus an entire data transformation (such as choosing top 8 athletes
by the number of gold medals) can be constructed using dot-driven development.

@@

Our language supports method calls and so some of the opertaions that are currently exposed as
member access could equally be provided as methods. For example, filtering could be written as 
$\qident{Games is}(\str{Rio (2016)})$. However, the fact that we can offer possible values when
filtering largely simplifies writing of the script for the most 
common case when the user is interested in one of the known values.

@@

\noindent
When grouping, we sum the number of gold and silver medals and concatenates distinct team names~\ball{1}. 
Then we sort the grouped data using two sorting keys \ball{2} -- first by the number of gold 
medals and then silver medals (within a group with the same number of gold medals).

@@

In type providers for F\#, renaming of columns could be encoded using methods with static parameters \cite{staticpar} by writing, 
for example, $\ident{g}.\qident{sum Gold as}\langl\str{Total Gold}\rangl()$. In F\#, the value of 
the static parameter (here, \str{Total Gold}) is passed to the type provider, which can use it to
generate the type signature of the method and the return type with member name according to the
value of the static parameter.

@@

\noindent
Operations that change the type of rows in the data set can be captured using row types~\cite{rowtypes}. 
Row types make it possible to statically track operations on records that add or remove fields
and so they can be used for the typing of operations such as $\qident{drop Year}$.
In addition, we need to annotate type with a form of typestate \cite{typestate} to restrict what operations are available. 
When dropping columns, we first access the \qident{drop columns} member, which sets the state to
a state where we can drop individual columns using \qident{drop $f$}. The \ident{then} member can 
then be used to complete the operation and choose another transformation.

@@

\begin{figure}
\vspace{-1em}
\begin{equation*}
\begin{array}{rcl}
  v &=& C(v)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(v)\; \lsep n \lsep s \lsep D \\
  e &=& C(e)\,\, \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(e)\;\, \lsep x \lsep v \lsep e.N \lsep \ldots\\
  E &=& C(E) \lsep \ident{series}\langl \tau_1, \tau_2 \rangl(E)\, \lsep E.N \\
    &|& \Pi_{f_1, \ldots, f_n}(E) \lsep \sigma_\varphi(E) \lsep \tau_{f_1, \ldots, f_n}(E) \lsep \Phi_{f, \rho_1/f_1, \ldots, \rho_m/f_m} (E) \\
    \\
 \tau &=& C \lsep \kvd{num} \lsep \kvd{string} \lsep \ident{series}\langl \tau_1, \tau_2 \rangl \lsep \ident{Query}\\
 l &=& \kvd{type}~C(x:\tau) = \overline{m} \\
 m &=& \kvd{member}~N:\tau= e
\end{array}
\end{equation*}
\vspace{-1.85em}

\[
\inference[(\emph{member})~]
{ L(C) = (\kvd{type}~C(x:\tau)= \ldots~\kvd{member}~N_i : \tau_i = e_i~\ldots), L'}
{ (C(v)).N_i \leadsto_L e_i[x \leftarrow v] }\\
\]
\vspace{-3.15em}

@@

The Foo calculus includes two special types. \ident{Query} is a type of data and queries 
constructed using the relational algebra. The type $\ident{series}\langl \tau_1, \tau_2 \rangl$ 
models a type-safe data series mapping keys of type $\tau_1$ to values of type $\tau_2$ that
can be used, for example, as input for a charting library. A series is a typed wrapper over a 
\ident{Query} value and the proofs in Section~\ref{sec:pivot-prop} show that a series obtained 
from the pivot type provider contains keys and values of matching~types.

@@

\noindent
The type checker initially knows the type of $\ident{olympics}$ is a class $C_1$ with member 
$\qident{group data}$ and it knows that the type of this member is $C_2$. However, it only needs to 
obtain full definition of $C_2$ when checking the member $\qident{by Athlete}$. Types of other 
members of $C_1$ remain unevaluated. This aspect of type
providers have been omitted in previous work \cite{fsdata,liteq}, but it is necessary for the pivot type provider.
The typing rules given are written as:
%
\begin{equation*}
L_1; \Gamma \vdash e : \tau; L_2
\end{equation*}

@@

\begin{figure}[t]
\vspace{-0.5em}
\[
\inference[(\emph{num})~]{}{L; \Gamma \vdash n : \kvd{num}; L }
\quad
\inference[(\emph{string})~]{}{L; \Gamma \vdash s : \kvd{string}; L}
\quad
\inference[(\emph{var})~]{}{L; \Gamma, x:\tau\vdash x:\tau; L}
\]
\vspace{-1.75em}

\[
\inference[(\emph{data})~]
  {L; \Gamma \vdash v_{i,j} : \tau;L \quad \tau \in \{ \kvd{num}, \kvd{string} \}}
  {L; \Gamma \vdash \{f_1\mapsto \vect{v_{1,1},\ldots,v_{1,r}}, \ldots, f_n\mapsto \vect{v_{n,1},\ldots,v_{n,r}}\} : \ident{Query};L }
\]
\vspace{-1.75em}

@@

\[
\inference[(\emph{sel})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \sigma_\varphi (e) : \ident{Query}; L_2}
\qquad
\inference[(\emph{group})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(e): \ident{Query}; L_2}
\]
\vspace{-1.75em}

\[
\inference[(\emph{series})~]
  {L_1; \Gamma\vdash e : \ident{Query}; L_2}
  {L_1; \Gamma\vdash \ident{series}\langl \tau_1, \tau_2 \rangl(e) : \ident{series}\langl \tau_1, \tau_2 \rangl; L_2}
\]
\vspace{-1.75em}

\[
\inference[(\emph{new})~]
  {L_1; \Gamma \vdash e : \tau, L_2 & L_2(C) = (\kvd{type}~C(x:\tau) = \ldots),L}
  {L_1; \Gamma \vdash C(e) : C; L_2 \cup L}
\]
\vspace{-1.75em}

\[
\inference[(\emph{member})~]
  {L_1; \Gamma \vdash e : C; L_2 \qquad L_2\cup L; \Gamma,x:\tau \vdash e_i : \tau_i; L_3 \\
   L_2(C)=(\kvd{type}~C(x:\tau) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..), L }
  {L_1; \Gamma \vdash e.N_i:\tau_i; L_3}
\]
\vspace{-0.5em}
\caption{Type-checking of Foo expressions with lazy context}
\label{fig:foo-typecheck}
\end{figure}

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{drop}(F) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\qquad\ball{1}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f \in \dom(F)~\textnormal{where}~C_f, L_f = \ident{drop}(F')\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f = C_f(\Pi_{\ident{dom}(F')}(x)) & \quad\textnormal{and}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)\qquad\ball{2}                 &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\\[0.5em]
\ident{sort}(F) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f) \}\qquad\ball{3}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\{f~|~F(f)=\kvd{num} \},~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{by $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{f \mapsto \ident{asc}})\\
\\[1em]
\multicolumn{2}{l}{ \ident{sort-and}(F,\vect{s_1,\ldots,s_n}) = C, \{ C \mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup L') \}\qquad\ball{4} }\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\hspace{-1em}\forall f\in\{f~|~F(f)=\kvd{num}, \,\nexists i.s_i=f'\mapsto\omega\wedge f'=f\}\\
\quad \qquad \kvd{member}~\qident{and $f$ descending} : C_f = C_f(x) &\quad C_f, L_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{desc}})\\
\quad \qquad \kvd{member}~\qident{and $f$ ascending} : C'_f = C'_f(x) &\quad C'_f, L'_f = \ident{sort-and}(F, \vect{s_1,..,s_n, f \mapsto \ident{asc}})\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(\tau_{s_1,\ldots,s_n}(x))\quad~~\ball{5}           &\textnormal{where}~C', L' = \ident{pivot}(F)\\
\end{array}
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\begin{array}{ll}
\ident{group}(F) = C, \{ C\mapsto (l, \bigcup L_f) \}\quad\ball{1} \\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)~\textnormal{where} \\
\quad \qquad \kvd{member}~\qident{by $f$} : C_f = C_f(x) &\quad C_f, L_f = \ident{agg}(F, f, \{ f \mapsto F(f) \}, \emptyset)~~\ball{2}\\
\\[0.5em]
\multicolumn{2}{l}{ \ident{agg}(F,f,G,S) = C, \{C\mapsto (l, \bigcup L_f \cup \bigcup L'_f \cup\bigcup L''_f\cup L'\cup L'') \}\quad\ball{3}}\\[0.25em]
~ l = \kvd{type}~C(x:\ident{Query}) = &\forall f\in\dom(F)\setminus\dom(S)\\
\quad \qquad \kvd{member}~\qident{sum $f$} : C'_f = C'_f(x) & \quad \textnormal{when}~F(f)=\kvd{num}\quad\ball{4} \\
\quad \qquad \kvd{member}~\qident{concat $f$} : C''_f = C''_f(x) & \quad \textnormal{when}~F(f)=\kvd{string}\quad\ball{5} \\
\quad \qquad \kvd{member}~\qident{count all} : C' = C'(x) & \quad \textnormal{when}~\ident{Count}\notin G\quad\ball{6} \\
\quad \qquad \kvd{member}~\qident{distinct $f$} : C_f = C_f(x)  \\
\quad \qquad \kvd{member}~\ident{then} : C'' = C''(\Phi_{f, \rho_1/f_1, \ldots, \rho_n/f_n}(x)) &\textnormal{where}~\{\rho_1/f_1,\ldots,\rho_n/f_n\}=S\quad\ball{7} \\[0.5em]
\qquad  \textnormal{where}\\[0.5em]
\multicolumn{2}{l}{
~\,\qquad \begin{array}{l}
 C_f, L_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num}\}, S\cup\{\ident{dist}~f/f\} )\\
 C'_f, L'_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{num} \}, S\cup\{\ident{sum}~f/f\})\\
 C''_f, L''_f = \ident{agg}(F,f, G\cup\{ f \mapsto \kvd{string} \}, S\cup\{\ident{conc}~f/f\})\\
 C', L' = \ident{agg}(F,f, G\cup\{ \ident{Count}\mapsto\kvd{int} \}, S\cup{\ident{count}/\ident{Count}})\\
 C'', L'' = \ident{pivot}(G)
\end{array} 
}
\end{array}
\end{equation*}

@@

\newpage

@@

\begin{figure}[t]
\vspace{-1em}
\begin{equation*}
\begin{array}{ll}
  \ident{pivot}(F, D) = C, \{ C \mapsto (l, L_1 \cup L_2 \cup \ldots  \}\quad \ball{1}\\[0.5em]
  ~ l = \kvd{type}~C(x:\ident{Query}) = \\
  \quad \qquad \kvd{member}~\qident{drop columns} : C_1 = C_1(x) &\textnormal{where}~C_1, L_1 = \ident{drop}(F, D)\\
  \quad \qquad \kvd{member}~\qident{filter data} : C_2 = C_2(x) &\textnormal{where}~C_2, L_2 = \ident{filter}(F, D)\\
  \quad \qquad (\ldots)\\
\\
\ident{drop}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\quad\ball{2}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{drop $f$} : C_f =  & \quad\textnormal{where}~ F' = \{ f'\mapsto\tau'\in F, f' \neq f\}\\
\quad \qquad \quad C_f(\Pi_{\ident{dom}(F')}(x)) &\quad\textnormal{and}~C_f, L_f = \ident{drop}(F', \Pi_{\ident{dom}(F')}(D))\quad\ball{3}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter}(F, D) = C, \{ C \mapsto (l, L' \cup \bigcup L_f) \}\\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall f \in \dom(F)~\\
\quad \qquad \kvd{member}~\qident{$f$ is} : C_f = C_f(x) & \quad\textnormal{where}~C_f, L_f = \ident{filter-val}(F, f, D)\quad\ball{4}\\
\quad \qquad \kvd{member}~\ident{then} : C' = C'(x)                 &\textnormal{where}~C', L' = \ident{pivot}(F, D)\\
\\
\ident{filter-val}(F, f, D) = C, \{ C \mapsto (l, \cup \bigcup L_v) \}&\textnormal{where}~D=\{ f\mapsto\vect{v_1, \ldots, v_n}, \ldots \}\quad\ball{5} \\[0.5em]
~ l = \kvd{type}~C(x:\ident{Query}) = & \forall v \in \{v_1, \ldots, v_n \} \\
\quad \qquad \kvd{member}~\qident{$\,v\,$} : C_v =  & \quad\textnormal{where}~C_v, L_v = \ident{filter}(F, \sigma_{\varphi_v}(D))\\
\quad \qquad \quad C_v(\sigma_{\varphi_v}(x)) &\quad\textnormal{and}~ \varphi_v (r) = r(f) = v \quad \ball{6}
\end{array}
\end{equation*}

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
\ident{olympics}.\qident{filter data}.\qident{Medal is}.\ident{Gold}.\qident{Team is}\\[0.25em]
\quad\begin{array}{l}
\rightarrow \qident{Czech Republic}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Barbora Spotakova}\\
\qquad \rightarrow \qident{David Kostelecky}\\
\qquad \rightarrow \qident{David Svoboda}\\
\end{array}\qquad
\begin{array}{l}
\rightarrow \ident{Mongolia}.\qident{Athlete is}\\[0.25em]
\qquad \rightarrow \qident{Badar-Uugan Enkhbat}\\
\qquad \rightarrow \qident{Tuvshinbayar Naidan}\\
\\
\end{array}
\end{array}
\end{equation*}
\caption{Subset of members provided by the filtering operation}
\label{fig:pivot-filter}
\end{figure}

@@

\begin{figure}[b]
\vspace{1em}
\begin{center}
\includegraphics[scale=0.42,trim=0mm 0mm 0mm 0mm,clip]{images/table2.png} % left bottom right top
\end{center}
\caption{Athletes by the number of medals over the entire history of Olympic games}
\label{fig:case-table}
\end{figure}

@@

~

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.35,trim=0mm 0mm 0mm 0mm,clip]{images/pivot.png} % left bottom right top
\end{center}
\caption{Spreadsheet-inspired live editor for the pivot type provider}
\label{fig:case-ed}
\end{figure}
